---
title: "Homework 2"
author: "Rachel Liu"
date: "2026-10-01"
output: github_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readxl)
```

## Problem 1
```{r}
subway_df = read_csv(
  "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
  show_col_types = FALSE
) |>
  select(
    Line,
    `Station Name`,
    `Station Latitude`,
    `Station Longitude`,
    Route1:Route11,
    `Entrance Type`,
    Entry,
    Vending,
    ADA
  ) |>
  rename_with(
    ~ str_replace_all(str_to_lower(.x), " ", "_")
  ) |>
  mutate(
    entry = entry == "YES",
    across(route1:route11, as.character)
  )

glimpse(subway_df)
dim(subway_df)
```
The cleaned subway dataset contains 1,868 observations and 19 variables. It includes subway line and station name, station latitude and longitude, routes served, entrance type, entry status, vending information, and ADA compliance. I retained the variables needed for the analysis, cleaned the variable names, converted entry from YES/NO to a logical variable, and converted the route variables to character type. The data are not fully tidy because the routes served are stored across multiple columns (route1 through route11) instead of in separate rows.

```{r}
subway_df |>
  distinct(line, station_name) |>
  nrow()

subway_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
  
subway_df |>
  filter(vending == "NO") |>
  summarize(
    proportion_allow_entry = mean(entry, na.rm = TRUE)
  )
```
There are 465 distinct subway stations in the dataset, of which 84 are ADA compliant. Among station entrances/exits without vending, approximately 37.7% allow entrance.

```{r}
subway_long = subway_df |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name",
    values_drop_na = TRUE
  ) |>
  mutate(
    route_number = as.integer(
      str_remove(route_number, "route")
    )
  )

glimpse(subway_long)

subway_long |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()

subway_long |>
  filter(route_name == "A", ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
```
After reshaping the route variables into a long format, 60 distinct stations serve the A train, and 17 of those stations are ADA compliant.

## Problem 2
```{r}
trash_file = "data/202509 Trash Wheel Collection Data.xlsx"

# Mr. Trash Wheel
mr_trash = read_excel(
  trash_file,
  sheet = "Mr. Trash Wheel",
  range = "A2:N709"
) |>
  rename_with(
    ~ .x |>
      str_to_lower() |>
      str_replace_all("[^a-z0-9]+", "_") |>
      str_remove("_$")
  ) |>
  mutate(
  month = str_trim(month),
  year = as.integer(year),
  dumpster = as.integer(dumpster),
  sports_balls = as.integer(round(sports_balls)),
  trash_wheel = "Mr. Trash Wheel"
)


# Professor Trash Wheel
professor_trash = read_excel(
  trash_file,
  sheet = "Professor Trash Wheel",
  range = "A2:M134"
) |>
  rename_with(
    ~ .x |>
      str_to_lower() |>
      str_replace_all("[^a-z0-9]+", "_") |>
      str_remove("_$")
  ) |>
  mutate(
      month = str_trim(month),
      year = as.integer(year),
      dumpster = as.integer(dumpster),
      trash_wheel = "Professor Trash Wheel"
    )

# Gwynnda Falls Trash Wheel
gwynnda_trash = read_excel(
  trash_file,
  sheet = "Gwynns Falls Trash Wheel",
  range = "A2:L351"
) |>
  rename_with(
    ~ .x |>
      str_to_lower() |>
      str_replace_all("[^a-z0-9]+", "_") |>
      str_remove("_$")
  ) |>
  mutate(
    month = str_trim(month),
    year = as.integer(year),
    dumpster = as.integer(dumpster),
    trash_wheel = "Gwynnda"
  )

trash_wheel_df = bind_rows(
  mr_trash,
  professor_trash,
  gwynnda_trash
)

glimpse(trash_wheel_df)
dim(trash_wheel_df)
```

```{r}
professor_total_weight = trash_wheel_df |>
  filter(trash_wheel == "Professor Trash Wheel") |>
  summarize(
    total_weight = sum(weight_tons, na.rm = TRUE)
  )

professor_total_weight

gwynnda_june_2022 = trash_wheel_df |>
  filter(
    trash_wheel == "Gwynnda",
    year == 2022,
    month == "June"
  ) |>
  summarize(
    total_cigarette_butts = sum(cigarette_butts, na.rm = TRUE)
  )

gwynnda_june_2022
```
The combined Trash Wheel dataset has 1,188 observations and 15 variables. The variables include dumpster number, month, year, collection date, trash weight and volume, several types of litter, and which Trash Wheel collected the trash. I cleaned each sheet separately, removed rows and columns that were not part of the actual data, cleaned the variable names, and rounded the sports ball counts for Mr. Trash Wheel to integers before combining the three datasets. Professor Trash Wheel collected 282.26 tons of trash in total, and Gwynnda collected 18,120 cigarette butts in June 2022.

## Problem 3

```{r}
mci_baseline = read_csv(
  "data/MCI_baseline.csv",
  skip = 1,
  na = c("", ".", "NA"),
  show_col_types = FALSE
)

mci_amyloid = read_csv(
  "data/mci_amyloid.csv",
  skip = 1,
  na = c("", ".", "NA"),
  show_col_types = FALSE
)
```

```{r}
baseline_clean = mci_baseline |>
  rename(
    study_id = ID,
    baseline_age = `Current Age`,
    sex = Sex,
    education = Education,
    apoe4 = apoe4,
    age_at_onset = `Age at onset`
  ) |>
  mutate(
    sex = factor(
      sex,
      levels = c(0, 1),
      labels = c("Female", "Male")
    ),
    apoe4 = factor(
      apoe4,
      levels = c(0, 1),
      labels = c("Non-carrier", "Carrier")
    )
  ) |>
  filter(
    is.na(age_at_onset) | age_at_onset > baseline_age
  )

glimpse(baseline_clean)
dim(baseline_clean)
```

```{r}
nrow(baseline_clean)

baseline_clean |>
  summarize(
    developed_mci = sum(!is.na(age_at_onset))
  )

baseline_clean |>
  summarize(
    average_baseline_age = mean(baseline_age, na.rm = TRUE)
  )

baseline_clean |>
  filter(sex == "Female") |>
  summarize(
    proportion_apoe4_carrier = mean(
      apoe4 == "Carrier",
      na.rm = TRUE
    )
  )
```

The cleaned baseline dataset contains 479 participants. Of these, 93 developed MCI during follow-up. The average baseline age was 65.03 years, and 30% of the women were APOE4 carriers. Sex and APOE4 carrier status were converted from numeric codes to categorical variables, and participants who already had MCI at baseline were removed.

```{r}
amyloid_clean = mci_amyloid |>
  rename(
    study_id = `Study ID`
  ) |>
  pivot_longer(
    cols = -study_id,
    names_to = "visit",
    values_to = "amyloid_42_40"
  ) |>
  mutate(
    time = parse_number(
      str_replace(visit, "Baseline", "0")
    )
  ) |>
  select(
    study_id,
    time,
    amyloid_42_40
  ) |>
  arrange(study_id, time)

glimpse(amyloid_clean)
dim(amyloid_clean)
```

The amyloid data were reshaped from wide to long format so that each row represents one participant at one measurement time. The cleaned amyloid dataset contains 2,435 observations and three variables: participant ID, time since baseline, and the amyloid 42/40 ratio.

```{r}
baseline_only = baseline_clean |>
  anti_join(
    amyloid_clean |>
      distinct(study_id),
    by = "study_id"
  )

baseline_only
```

```{r}
amyloid_only = amyloid_clean |>
  distinct(study_id) |>
  anti_join(
    baseline_clean |>
      distinct(study_id),
    by = "study_id"
  )

amyloid_only
```

There are 8 participants who appear only in the baseline dataset and 16 participants who appear only in the amyloid dataset.

```{r}
mci_combined = amyloid_clean |>
  inner_join(
    baseline_clean,
    by = "study_id"
  )

glimpse(mci_combined)
dim(mci_combined)

n_distinct(mci_combined$study_id)
```

```{r}
write_csv(
  mci_combined,
  "data/mci_combined.csv"
)
```

After keeping only participants who appear in both datasets, the final dataset contains 471 participants and 2,355 longitudinal observations. It includes the amyloid measurements and measurement time together with baseline age, sex, education, APOE4 carrier status, and age at MCI onset.