---
title: "Homework 2"
author: "Rachel Liu"
date: "2026-10-01"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(tidyverse)
```
## Problem 1

```{r}
subway_df = read_csv(
  "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
  show_col_types = FALSE
) |>
  select(
    Line,
    `Station Name`,
    `Station Latitude`,
    `Station Longitude`,
    Route1:Route11,
    `Entrance Type`,
    Entry,
    Vending,
    ADA
  ) |>
  rename_with(
    ~ str_replace_all(str_to_lower(.x), " ", "_")
  ) |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    ),
    across(route1:route11, as.character)
  )

glimpse(subway_df)
dim(subway_df)
```
The cleaned subway dataset contains 1,868 observations and 19 variables. It includes subway line and station name, station latitude and longitude, routes served, entrance type, entry status, vending information, and ADA compliance. I retained the variables needed for the analysis, cleaned the variable names, converted entry from YES/NO to a logical variable, and converted the route variables to character type. The data are not fully tidy because the routes served are stored across multiple columns (route1 through route11) instead of in separate rows.

```{r}
# Number of distinct stations
subway_df |>
  distinct(line, station_name) |>
  nrow()


# Number of ADA-compliant stations
subway_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
  
# Proportion of entrances/exits without vending that allow entry
subway_df |>
  filter(vending == "NO") |>
  summarize(
    proportion_allow_entry = mean(entry, na.rm = TRUE)
  )
```
There are 465 distinct subway stations in the dataset, of which 84 are ADA compliant. Among station entrances/exits without vending, approximately 37.7% allow entrance.

```{r}
subway_long = subway_df |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name",
    values_drop_na = TRUE
  ) |>
  mutate(
    route_number = as.integer(
      str_remove(route_number, "route")
    )
  )

glimpse(subway_long)

# Distinct stations serving the A train
subway_long |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()

# ADA-compliant stations serving the A train
subway_long |>
  filter(route_name == "A", ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
```
After reshaping the route variables into a long format, 60 distinct stations serve the A train, and 17 of those stations are ADA compliant.