---
title: "Homework 2"
author: "Rachel Liu"
date: "2026-10-01"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(tidyverse)
library(readxl)
```

## Problem 1
```{r}
subway_df = read_csv(
  "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
  show_col_types = FALSE
) |>
  select(
    Line,
    `Station Name`,
    `Station Latitude`,
    `Station Longitude`,
    Route1:Route11,
    `Entrance Type`,
    Entry,
    Vending,
    ADA
  ) |>
  rename_with(
    ~ str_replace_all(str_to_lower(.x), " ", "_")
  ) |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    ),
    across(route1:route11, as.character)
  )

glimpse(subway_df)
dim(subway_df)
```
The cleaned subway dataset contains 1,868 observations and 19 variables. It includes subway line and station name, station latitude and longitude, routes served, entrance type, entry status, vending information, and ADA compliance. I retained the variables needed for the analysis, cleaned the variable names, converted entry from YES/NO to a logical variable, and converted the route variables to character type. The data are not fully tidy because the routes served are stored across multiple columns (route1 through route11) instead of in separate rows.

```{r}
# Number of distinct stations
subway_df |>
  distinct(line, station_name) |>
  nrow()


# Number of ADA-compliant stations
subway_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
  
# Proportion of entrances/exits without vending that allow entry
subway_df |>
  filter(vending == "NO") |>
  summarize(
    proportion_allow_entry = mean(entry, na.rm = TRUE)
  )
```
There are 465 distinct subway stations in the dataset, of which 84 are ADA compliant. Among station entrances/exits without vending, approximately 37.7% allow entrance.

```{r}
subway_long = subway_df |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name",
    values_drop_na = TRUE
  ) |>
  mutate(
    route_number = as.integer(
      str_remove(route_number, "route")
    )
  )

glimpse(subway_long)

# Distinct stations serving the A train
subway_long |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()

# ADA-compliant stations serving the A train
subway_long |>
  filter(route_name == "A", ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
```
After reshaping the route variables into a long format, 60 distinct stations serve the A train, and 17 of those stations are ADA compliant.

## Problem 2
```{r}
trash_file = "data/202509 Trash Wheel Collection Data.xlsx"

# Mr. Trash Wheel
mr_trash = read_excel(
  trash_file,
  sheet = "Mr. Trash Wheel",
  range = "A2:N709"
) |>
  rename_with(
    ~ .x |>
      str_to_lower() |>
      str_replace_all("[^a-z0-9]+", "_") |>
      str_remove("_$")
  ) |>
  mutate(
  month = str_trim(month),
  year = as.integer(year),
  dumpster = as.integer(dumpster),
  sports_balls = as.integer(round(sports_balls)),
  trash_wheel = "Mr. Trash Wheel"
)


# Professor Trash Wheel
professor_trash = read_excel(
  trash_file,
  sheet = "Professor Trash Wheel",
  range = "A2:M134"
) |>
  rename_with(
    ~ .x |>
      str_to_lower() |>
      str_replace_all("[^a-z0-9]+", "_") |>
      str_remove("_$")
  ) |>
  mutate(
      month = str_trim(month),
      year = as.integer(year),
      dumpster = as.integer(dumpster),
      trash_wheel = "Professor Trash Wheel"
    )


# Gwynnda / Gwynns Falls Trash Wheel
gwynnda_trash = read_excel(
  trash_file,
  sheet = "Gwynns Falls Trash Wheel",
  range = "A2:L351"
) |>
  rename_with(
    ~ .x |>
      str_to_lower() |>
      str_replace_all("[^a-z0-9]+", "_") |>
      str_remove("_$")
  ) |>
  mutate(
    month = str_trim(month),
    year = as.integer(year),
    dumpster = as.integer(dumpster),
    trash_wheel = "Gwynnda"
  )

trash_wheel_df = bind_rows(
  mr_trash,
  professor_trash,
  gwynnda_trash
)

glimpse(trash_wheel_df)
dim(trash_wheel_df)
```

```{r}
# Total weight collected by Professor Trash Wheel
professor_total_weight = trash_wheel_df |>
  filter(trash_wheel == "Professor Trash Wheel") |>
  summarize(
    total_weight = sum(weight_tons, na.rm = TRUE)
  )

professor_total_weight


# Cigarette butts collected by Gwynnda in June 2022
gwynnda_june_2022 = trash_wheel_df |>
  filter(
    trash_wheel == "Gwynnda",
    year == 2022,
    month == "June"
  ) |>
  summarize(
    total_cigarette_butts = sum(cigarette_butts, na.rm = TRUE)
  )

gwynnda_june_2022
```
The combined Trash Wheel dataset has 1,188 observations and 15 variables. The variables include dumpster number, month, year, collection date, trash weight and volume, several types of litter, and which Trash Wheel collected the trash. I cleaned each sheet separately, removed rows and columns that were not part of the actual data, cleaned the variable names, and rounded the sports ball counts for Mr. Trash Wheel to integers before combining the three datasets. Professor Trash Wheel collected 282.26 tons of trash in total, and Gwynnda collected 18,120 cigarette butts in June 2022.

## Problem 3
```{r}
mci_baseline = read_csv(
  "data/MCI_baseline.csv",
  skip = 1,
  show_col_types = FALSE
)

mci_amyloid = read_csv(
  "data/mci_amyloid.csv",
  skip = 1,
  show_col_types = FALSE
)

glimpse(mci_baseline)
names(mci_baseline)

glimpse(mci_amyloid)
names(mci_amyloid)
```

```{r}
baseline_clean = mci_baseline |>
  rename(
    study_id = ID,
    baseline_age = `Current Age`,
    sex = Sex,
    education = Education,
    apoe4 = apoe4,
    age_at_onset = `Age at onset`
  ) |>
  mutate(
    age_at_onset = as.numeric(na_if(age_at_onset, ".")),
    sex = factor(
      sex,
      levels = c(0, 1),
      labels = c("Female", "Male")
    ),
    apoe4 = factor(
      apoe4,
      levels = c(0, 1),
      labels = c("Non-carrier", "Carrier")
    )
  ) |>
# Keep only participants without MCI at baseline
  filter(
    is.na(age_at_onset) | age_at_onset > baseline_age
  )

glimpse(baseline_clean)
dim(baseline_clean)

# Number of participants recruited
nrow(baseline_clean)

# Number who developed MCI during follow-up
baseline_clean |>
  summarize(
    developed_mci = sum(!is.na(age_at_onset))
  )

# Average baseline age
baseline_clean |>
  summarize(
    average_baseline_age = mean(baseline_age, na.rm = TRUE)
  )

# Proportion of women who are APOE4 carriers
baseline_clean |>
  filter(sex == "Female") |>
  summarize(
    proportion_apoe4_carrier = mean(apoe4 == "Carrier", na.rm = TRUE)
  )

amyloid_clean = mci_amyloid |>
  rename(
    study_id = `Study ID`
  ) |>
  mutate(
    across(
      -study_id,
      ~ as.numeric(na_if(.x, "."))
    )
  ) |>
  pivot_longer(
    cols = -study_id,
    names_to = "visit",
    values_to = "amyloid_42_40"
  ) |>
  mutate(
    time = case_when(
      visit == "Baseline" ~ 0,
      TRUE ~ parse_number(visit)
    )
  ) |>
  select(
    study_id,
    time,
    amyloid_42_40
  ) |>
  arrange(study_id, time)

glimpse(amyloid_clean)
dim(amyloid_clean)
# Participants only in baseline dataset
baseline_only = baseline_clean |>
  anti_join(
    amyloid_clean |> distinct(study_id),
    by = "study_id"
  )

baseline_only


# Participants only in amyloid dataset
amyloid_only = amyloid_clean |>
  distinct(study_id) |>
  anti_join(
    baseline_clean |> distinct(study_id),
    by = "study_id"
  )

amyloid_only
mci_combined = amyloid_clean |>
  inner_join(
    baseline_clean,
    by = "study_id"
  )

glimpse(mci_combined)
dim(mci_combined)
write_csv(
  mci_combined,
  "data/mci_combined.csv"
)
```
After cleaning the baseline data and removing participants who already had MCI at baseline, 479 participants remained. Of these, 93 developed MCI during follow-up. The average baseline age was 65.03 years, and 30% of women were APOE4 carriers. The amyloid data were reshaped into long format, with one row for each participant and measurement time. There were 8 participants found only in the baseline dataset and 16 found only in the amyloid dataset. After keeping participants present in both datasets, the final dataset contained 471 participants and 2,355 observations.